Separating Use and Reuse to Improve Both

### 2017-06-06 FACS/Intro.tex
Revision

\section{Introduction}\label{sec:intro}
\saveSpace\saveSpace
In mainstream OO languages like Java, C++ or C\# subclassing 
implies subtyping. For example, in Java a subclass definition, such as 
\Q@class A extends B {}@
\noindent does two things at the same time:
it {\bf inherits} code from \lstinline{B}; and it {\bf creates
a subtype} of \emph{B}. Therefore in a language like Java 
a subclass is \emph{always} a subtype of the extended class.

@@

What the this-leaking problem shows is that adopting a more flexible
nominally typed OO model where subclassing does not imply subtyping is
not trivial: a more substantial change in the language design is
necessary.  In essence we believe that, in languages like Java, classes do too many
things at once. In particular they act both as units of \emph{use} and
\emph{reuse}: classes can be \emph{use}d as types and can be instantiated;
classes can also be subclassed to provide \emph{reuse} of code.
We are aware of at least 3 independently designed research
languages that address the this-leaking problem:
\begin{itemize}
\item In {\bf Delta-Trait (DT)}~\cite{Bettini:2010:ISP:1774088.1774530,BETTINI2013521,Bettini2015282}
each  construct has a single responsability: classes instantiate objects,
interfaces induce types, records express state and traits are reuse units.
\item {\bf Package Templates (PT)}~\cite{KrogdahlMS09,DBLP:journals/taosd/AxelsenSKM12,DBLP:conf/gpce/AxelsenK12}
are an extension of (full) Java where new packages can be ``synthesized'' by mixing
and integrating code templates. As an extension of Java, PT allows but not require
separation of inheritance and subtyping.
\item {\bf
    DeepFjig}~\cite{deep,servetto2014meta,fjig} is
a module composition language where the main idea is that
nested classes with the same name are recursivelly composed.
\end{itemize}
This paper aims at showing a simple language design, called \name.
Leveraging on traits, in this work we aim to synthesize the best ideas
of those 3 very different designs, while at the same time coming up with a simpler and
improved design for separating subclassing from subtyping.
The keys ideas in \name are to divide between code designed for
\textbf{USE} and code designed for \textbf{REUSE}, and a novel
approach to state in traits that avoids the complexities introduced by
fields and their initialization.
In \name there are two separate concepts: classes
and traits~\cite{Traits:ECOOP2003}. Classes are meant for code use, and cannot be used
for reuse. In some sense classes in \name are like final classes in
Java. Traits are meant for code reuse and multiple traits can be
composed to form a class which can then be instantiated. Traits 
cannot be instantiated (or used) directly. Such design allows the
subtyping and code reuse to be treated separately, which in turn
brings several benefits in terms of flexibility and code reuse.

We focus on an example-driven presentation to illustrate how to
improve use and reuse. Most of the hard technical aspects of the
semantics have been studied in previous 
work~\cite{Bettini:2010:ISP:1774088.1774530,BETTINI2013521,Bettini2015282,KrogdahlMS09,DBLP:journals/taosd/AxelsenSKM12,DBLP:conf/gpce/AxelsenK12,deep,servetto2014meta,fjig},
and the design of \name synthesises some of the previously studied
concepts. We briefly sketch the key ideas of the formalization of
\name in Section~\ref{sec:formal}.
The language design ideas have been implemented in the 42 language, which supports all
the examples we show in the paper, and is available at:

@@

\begin{itemize}
\item We identify the {\bf this-leaking} problem and argue why it
  makes the separation of subclassing and subtyping difficult.
\item We synthesize the key ideas of previous designs that solve the
  this-leaking problem into {\bf a novel and
  simple language design}. This language is the logic core of the language 42, and 
  all the examples in the paper can be directly encoded as valid 42 programs.

>>>>

\section{Introduction}\label{sec:intro}
\saveSpace\saveSpace
In mainstream OO languages like Java, C++ or C\# subclassing 
implies subtyping. For example, a Java subclass definition, such as 
\Q@class A extends B {}@
\noindent does two things at the same time:
it {\bf inherits} code from \lstinline{B}; and it {\bf creates
a subtype} of \emph{B}. Therefore in a language like Java 
a subclass is \emph{always} a subtype of the extended class.

@@

What the this-leaking problem shows is that adopting a more flexible
nominally typed OO model where subclassing does not imply subtyping is
not trivial: a more substantial change in the language design is
necessary.  In essence we believe that, in languages like Java, classes do too many
things at once. In particular they act both as units of \emph{use} and
\emph{reuse}: classes can be \emph{use}d as types and can be instantiated;
classes can also be subclassed to provide \emph{reuse} of code.
We are aware of at least 3 independently designed research
languages that address the this-leaking problem:
\begin{itemize}
\item In {\bf Delta-Trait (DT)}~\cite{Bettini:2010:ISP:1774088.1774530,BETTINI2013521,Bettini2015282}
each  construct has a single responsability: classes instantiate objects,
interfaces induce types, records express state and traits are reuse units.
\item {\bf Package Templates (PT)}~\cite{KrogdahlMS09,DBLP:journals/taosd/AxelsenSKM12,DBLP:conf/gpce/AxelsenK12}
are an extension of (full) Java where new packages can be ``synthesized'' by mixing
and integrating code templates. As an extension of Java, PT allows but not require
separation of inheritance and subtyping.
\item {\bf
    DeepFjig}~\cite{deep,servetto2014meta,fjig} is
a module composition language where the main idea is that
nested classes with the same name are recursivelly composed.
\end{itemize}
This paper aims at showing a simple language design, called \name,
which addresses the this-leaking problem and decouples subtyping from inheritance.
Leveraging on traits, in this work we aim to synthesize the best ideas
of those 3 very different designs, while at the same time coming up with a simpler and
improved design for separating subclassing from subtyping.
The keys ideas in \name are to divide between code designed for
\textbf{use} and code designed for \textbf{reuse}, and a novel
approach to state in traits that avoids the complexities introduced by
fields and their initialization in previous approaches.
In \name there are two separate concepts: classes
and traits~\cite{Traits:ECOOP2003}. Classes are meant for code use, and cannot be used
for reuse. In some sense classes in \name are like final classes in
Java. Traits are meant for code reuse and multiple traits can be
composed to form a class which can then be instantiated. Traits 
cannot be instantiated (or used) directly. Such design allows the
subtyping and code reuse to be treated separately, which in turn
brings several benefits in terms of flexibility and code reuse.

We focus on an example-driven presentation to illustrate how to
improve use and reuse. Most of the hard technical aspects of the
semantics have been studied in previous 
work~\cite{Bettini:2010:ISP:1774088.1774530,BETTINI2013521,Bettini2015282,KrogdahlMS09,DBLP:journals/taosd/AxelsenSKM12,DBLP:conf/gpce/AxelsenK12,deep,servetto2014meta,fjig},
and the design of \name synthesises some of the previously studied
concepts. We briefly sketch the key ideas of the formalization of
\name in Section~\ref{sec:formal}.
The language design ideas have been implemented in the {\bf 42 language}, which supports all
the examples we show in the paper, and is available at:

@@

\begin{itemize}
\item We identify the {\bf this-leaking} problem and argue why it
  makes the separation of subclassing and subtyping difficult.
\item We synthesize the key ideas of previous designs that solve the
  this-leaking problem into {\bf a novel and
  simple language design}. This language is the logic core of the language 42, and 
  all the examples in the paper can be encoded as valid 42 programs.

### 2017-06-06 FACS/paper.tex
Section 3
Revision

\begin{abstract}
\saveSpace\saveSpace\saveSpace
In most OO languages subclassing/inheritance implies
subtyping. This is considered by many a conceptual design error, but it
seems required for technical reasons, due to what we call the
\emph{this-leaking problem}. This problem shows that separating
inheritance from subtyping is non-trivial and requires a significant
departure from the OO models in existing mainstream OO languages.
We are aware of at least 3 independently designed research languages 
addressing this limitation: \emph{Delta-Trait}, \emph{Package Template}, \emph{DeepFjig}.
The goal of this paper is to synthesize and improve on
the main ideas of those very different designs into a nominally typed
minimalist language, called \name, which is quite
natural to OO programmers.
By making our type system distinguish between code-use and code-reuse
we can separate inheritance and subtyping, while avoiding 
redundant abstract declarations required in Delta-Trait and
DeepFjig. At the same time self construction,
binary methods and recursive types are also supported.
Moreover, we provide a novel and elegant solution to uniformly
handle behavior and state within trait composition.
These ideas have been implemented in the 42 language, 
that supports all the examples we show in the paper.
\saveSpace\saveSpace\saveSpace

>>>>

\begin{abstract}
\saveSpace\saveSpace\saveSpace
In most OO languages subclassing/inheritance implies
subtyping. This is considered by many a conceptual design error, but it
seems required for technical reasons, due to what we call the
\emph{this-leaking problem}. This problem shows that separating
inheritance from subtyping is non-trivial and requires a significant
departure from the OO models in existing mainstream OO languages.
We are aware of at least 3 independently designed research languages 
addressing this limitation: \emph{Delta-Trait}, \emph{Package Template}, \emph{DeepFjig}.
The goal of this paper is to synthesize and improve on
the main ideas of those very different designs into a nominally typed
minimalist language, called \name, which is quite
natural to OO programmers.
By making our type system distinguish between code-use and code-reuse
we can separate inheritance and subtyping, while avoiding 
redundant abstract declarations required in Delta-Trait and
DeepFjig. At the same time \emph{self construction},
\emph{binary methods} and \emph{recursive types} are also supported.
Moreover, we provide a novel and elegant solution to uniformly
handle behavior and state within trait composition.
These ideas have been implemented in the 42 language, 
which supports all the examples we show in the paper.
\saveSpace\saveSpace\saveSpace
